Learning constraints in tabular data

### 2016-05-19 Paper/ECAI/tabular.tex
switch to IEEEtran for DSAA
more on the formalisation

\documentclass{sig-alternate-05-2015}

@@

%\usepackage{amsthm} % incompatible with ACM
%\theoremstyle{definition}
%\newtheorem{definition}{Definition}
\newdef{definition}{Definition} % ACM specific

@@

  \begin{subfigure}{.70\textwidth}
  \begin{center}
    \includegraphics[width=1\textwidth]{figures/Demo.png}
  \end{center}
  \vspace{-10pt}
  \caption{Example spreadsheet (black words and numbers only). Green borders indicate headerless tables; the blue elements are added to clarify some constraints and relations. \tias{TODO: headers not in green boxes} \sergey{show blocks? orientation only over blocks, not over tables?}\tias{Use specific constraint names, e.g. 'SUM-row' above Total in T1}}
  \label{fig:main_example}
\end{subfigure}
\hfill
\begin{subfigure}{.30\textwidth}
  {\small
    \begin{align*}
      &~\ecseries{\range{T_{1}}{\rangeall}{1}} \\
%
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
%
      &~\ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecaggc{\textit{AVERAGE}}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MAX}}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MIN}}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggif{\textit{SUM}}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      &~\ecaggif{\textit{MAX}}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      &~\eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
      &~\range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}}
  \vspace{-10pt}
  \caption{Constraints that are satisfied in the case study (Figure on the left) and expected to be found by our implementation. Non-functional constraint are omitted \tias{Add them too}.}
  \label{fig:sol_example}
\end{subfigure}
\end{figure*}

@@

@@

A row or columns is \textbf{type-consistent} if all cells in that row or column are of the same base type, that is, numeric or textual. We will use notation $T[a,{:}]$ to refer to the $a$-th row of table $T$, and similarly $T[{:},a]$ for the $a$-th column. For example in Figure~\ref{fig:main_example}, $T_1[1,:] = [1,2,3,4]$ and $T_5[:,1] = ['$Diana Coolen$', 5]$. The latter is not type-consistent while the former is.
We will write \textbf{vector} to refer to a single row or column when its orientation does not matter.

The most important concept is that of a \textbf{block}. A block has three properties: 1) contains only entire rows or entire columns of a single headerless table 2) it is contiguous and 3) it is type-consistent.
If it contains only rows we say it has \textit{row-orientation}, if only columns, \textit{column-orientation}. The rows or columns have to be contiguous in the original table meaning that they must visually form a block in the table; and each of the rows/columns has to be of the same type.

Because of these properties, we can use the following notation to refer to blocks: $B = T[\rangeto{a}{b},:]$ for a row-oriented block containing rows $a$ to $b$ in table $T$; and similarly $B = T[{:},\rangeto{a}{b}]$ for a column-oriented block.
%
The \textit{size} of a block is the number of vectors in the block, e.g. $b-a+1$, the number of rows or columns. The \textit{length} of a block is the length of its vectors; as all vectors are from the same table, they always have the same length. %We call a group $G$ \textit{numeric} (\textit{textual}, etc), written as \textit{numeric(G)}, if all its vectors contain numeric (textual, etc) elements.

@@

\tias{TODO: update everything to speak about blocks instead of groups}

\subsection{Constraint templates}
Our method will extract the applicable constraints (including functions) in a set of tables, based on \textit{constraint templates}.
%
A \template is a triple \textit{(\CName, \CSignature, \CFunction)}:
%Let us elaborate on this:
\begin{itemize}
\item
\textit{\CName}  specifies the syntactic form of the constraint $c(G_1, ...,G_n)$, that is, the name of the constraint $c$ together
with $n$ arguments $G_i$. Thus a constraint $c$ is viewed as a relation or predicate of arity $n$ in first order logic. Note that a function $G_r=f(G_1,...,G_n)$ can be represented with the $n+1$-ary predicate $cf(G_r,G_1,...,G_n)$.
% the list of its variables $v_1,\dots,v_n$.  In ILP terminology, it is known as vocabulary.

\item the \textit{\CSignature} defines the properties that the arguments of the predicate must satisfy. That is, the required types of the arguments, together
with possible further restrictions, e.g., on their length. For example, the signature can restrict the types of the arguments to be groups of integers and require that the length of all arguments be equal. In terms of logical and relational learning \cite{luc_book}, such a \CSignature is known as the {\em bias} of the learner, it specifies when the constraint is well-formed.

\item \textit{\CFunction} is the actual definition of the constraint that specifies when the constraint holds. Given an assignments to its arguments, it can be used to verify whether the constraint is satisfied or not. %holds, in practice this will be a function that can be called when
%the arguments are full instantiated to decide whether the constraint is satisfied or not.
In logical and relational learning this is known as the background knowledge, which provides the definition of the predicate.
\end{itemize}

\paragraph{Example}
Several example constraints are illustrated in Table \ref{table:constraints}.  For example, the constraint template \textit{rank} has
\begin{itemize}
  \item \CName: \ecrank{G_r}{G_x}, where $G_r$ and $G_x$ are the arguments;
\item \CSignature: the two arguments $G_r$ and $G_x$ must be numeric groups and of size one; furthermore, the group $G_r$ must be integer (the rank) and the length of groups $G_x$ and $G_r$ must be identical, i.e., $\plength(G_r) = \plength(G_x)$;
\item \CFunction: \ecrank{G_r}{G_x}, whenever each value of $G_r$ is the rank of the corresponding value in $G_x$.
% is satisfied in the mapping $\{ v_x \mapsto g_x, v_y \mapsto g_y \}$ satisfies the constraint iff $g_x \in G_x, g_y \in G_y$ ($G_x \in \groups, G_y \in \groups$) and each value in $g_y$ is the rank of the corresponding value in $g_x$ (possibly with ties).

\tias{Maybe better row-sum so that actual 'blocks' is clear instead of just vectors? (and then use same row-sum for all subsequent examples}
\end{itemize}

@@

%\sergey{moved from the formalization BEGIN}
%l 1) a set of cells, and 2) a set of constraints over those cells.  Notice that the constraints specify the types (or the domains) that one encounters in traditional CSPs.  Notice also, that as in traditional CSPs, the CSP problem defines the variables (the cells) of the problem as well as the constraint that should hold amongst them. The task of CSP-solvers is then to find an assignment to the variables that satisfies all the constraints.  The distinction between variables (cells) and their assignments (their values) is important here.  So, when we speak about cells, vectors and groups, we refer to the variables; when we talk about values, assignments or instances, we refer to the values these cells take in a particular "instantiated" table.

>>>>

\documentclass{IEEEtran}

@@

\usepackage{amsthm} % incompatible with ACM
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
%\newdef{definition}{Definition} % ACM specific

@@

  \begin{subfigure}{.70\textwidth}
  \begin{center}
    \includegraphics[width=1\textwidth]{figures/Demo.png}
  \end{center}
  \vspace{-10pt}
  \caption{Example spreadsheet (black words and numbers only). Green borders indicate headerless tables; the blue elements are added to clarify some constraints and relations. \tias{TODO: headers not in green boxes} \sergey{show blocks? orientation only over blocks, not over tables?}\tias{Use specific constraint names, e.g. 'SUM-row' above Total in T1}}
  \label{fig:main_example}
\end{subfigure}
\hfill
\begin{subfigure}{.30\textwidth}
  {\small
    \begin{align*}
      &~\ecseries{\range{T_{1}}{\rangeall}{1}} \\
%
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
%
      &~\ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecaggc{\textit{AVERAGE}}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MAX}}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MIN}}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggif{\textit{SUM}}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      &~\ecaggif{\textit{MAX}}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      &~\eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
      &~\range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}}
  \vspace{-10pt}
  \caption{Constraints present in the running example (left) \tias{Add ALL, including non-functional and duplicates}.}
  \label{fig:sol_example}
\end{subfigure}
\end{figure*}

@@

\tias{section description (add prob def here?)}

@@

A row or columns is \textbf{type-consistent} if all cells in that row or column are of the same base type, that is, numeric or textual. We will use notation $T[a,{:}]$ to refer to the $a$-th row of table $T$, and similarly $T[{:},a]$ for the $a$-th column. For example in Figure~\ref{fig:main_example}, $T_1[1,:] = [1,2,3,4]$ and $T_5[:,1] = ['$Diana Coolen$', 5]$. The latter is not type-consistent while the former is.
%We will write \textbf{vector} to refer to a single row or column when its orientation does not matter.

The most important concept is that of a \textbf{block}. A block has to satisfy three conditions: 1) it contains only entire rows or entire columns of a single headerless table 2) it is contiguous and 3) it is type-consistent.
If it contains only rows we say it has \textit{row-orientation}, if only columns, \textit{column-orientation}. The rows or columns have to be contiguous in the original table meaning that they must visually form a block in the table; and each of the rows/columns has to be of the same type. We will refer to the \textit{vectors} of a block when we wish to refer to its rows/columns independent of their orientation.

Because of these properties, we can use the following notation to refer to blocks: $B = T[\rangeto{a}{b},:]$ for a row-oriented block containing rows $a$ to $b$ in table $T$; and similarly $B = T[{:},\rangeto{a}{b}]$ for a column-oriented block.

A block has the following properties:
\begin{itemize}
\item \textit{type}: a block has to be type-consistent, so it has a single type
\item \textit{table}: the table that this block belongs to
\item \textit{orientation}: either row-oriented or column-oriented
\item \textit{size}: the size of a block is the number of vectors in it, e.g. $b-a+1$, the number of rows or columns.
\item \textit{length}: the length of its vectors; as all vectors are from the same table, they always have the same length.
\item \textit{rows}: this is orientation specific, for row-based blocks this is the size, for column-based blocks the length
\item \textit{columns}: also orientation specific, for row-based blocks this is the length, for column-based blocks the size
\item \tias{What is missing is 'discrete'}
\end{itemize}

%We call a group $G$ \textit{numeric} (\textit{textual}, etc), written as \textit{numeric(G)}, if all its vectors contain numeric (textual, etc) elements.

@@

\subsection{Constraint templates}
The goal is to learn constraints over blocks in the data. The knowledge needed to learn a constraint is expressed through {\template}s.
%
A \template $s$ is a triple $s=\textit{(\CName, \CSignature, \CFunction)}$:
%Let us elaborate on this:
\begin{itemize}
\item
\textit{\CName}  specifies the syntactic form of the constraint $c(G_1, ...,G_n)$, that is, the name of the constraint $c$ together
with $n$ abstract arguments $G_i$. Thus a constraint $c$ is viewed as a relation or predicate of arity $n$ in first order logic. Note that a function $G_r=f(G_1,...,G_n)$ can be represented with the $(n{+}1)$-ary predicate $c_f(G_r,G_1,...,G_n)$. Each argument will have to be instantiated with a block.
% the list of its variables $v_1,\dots,v_n$.  In ILP terminology, it is known as vocabulary.

\item the \textit{\CSignature} defines the properties that the arguments of the predicate must satisfy, that is, the blocks. This can be any property of individual blocks, such as type, length, size and orientation, as well as relations between properties of arguments, for example that the corresponding blocks must belong to the same table, or have equal type or length. In terms of logical and relational learning \cite{luc_book}, such a \CSignature is known as the {\em bias} of the learner, it specifies when the constraint its arguments are well-formed.

\item \textit{\CFunction} is the actual definition of the constraint that specifies when the constraint holds. Given an assignments of blocks to its arguments, it can be used to verify whether the constraint is satisfied or not. %holds, in practice this will be a function that can be called when
%the arguments are full instantiated to decide whether the constraint is satisfied or not.
In logical and relational learning this is known as the background knowledge. %, which provides the definition of the predicate.
\end{itemize}

\begin{example}
Several example constraint templates are illustrated in Table \ref{table:constraints}. An non-trivial example is the constraint template for the row-based sum:

\begin{itemize}
  \item \CName: $\ecsumr{\sg_r}{\mathbf{\sg_x}}$, where $\sg_r$ and $\mathbf{\sg_x}$ are the arguments;

  \item \CSignature: $\sg_r$ has to be a single vector ($size=1$) while $\mathbf{\sg_x}$ can be a block ($size>=1$), which can be derived from the use of a normal or \textbf{bold} font. The two blocks have to be numeric. This constraint is orientation-specific, so it requires that the number of rows in $\mathbf{\sg_x}$ equals the length of $\sg_r$. While not strictly needed, we also add to the bias of this template that the number of columns to sum over is larger than $2$ as a block with two columns will also be captured by the $\eccalc{\sg_r}{\sg_1 + \sg_2}$ constraint;

  \item \CFunction: each value in the vector $\sg_{r}$ is obtained by summing over the corresponding row in $\mathbf{\sg_x}$.
\end{itemize}
\end{example}

@@
